---
title: "Case Study_2"
author: "Kumar Ramasundaran, Julius Mwangi"
date: "July 29, 2018"
output: html_document
---

##  Pumps Predictive Analytics and Anomaly Detection

### INTRODUCTION

Pump predictive analytics is a process designed to predict pump failure using analysis of several failure variables. Why is it necessary to predict pump failure? First, pumps are used to push liquids, gasses, and other light solids through a mechanical process from one location to another.^[1] They simplify the shifting processes and improve on equipment operational efficiencies. They perform essential functions and their failure may result loss of revenue, more equipment collapsing, escalated labor costs due to stalled equipments etc. Secondly, In some cases, the pumps are pumping corrosive and environmentally hazardous materials. When they fail, they create unsafe working conditions. We want to predict pump functionality so we can eliminate and minimize such hazardous situations. To prevent catastrophic failures, Plant Managers, once a month or quarterly send their technicians, or hire contractors, to inspect these pumps. In this situation, the technicians/contractors risk exposure to hazardous materials. It is critical that these pieces of equipment are kept in functioning order as much as possible. This is where prediction comes in. The Plant Manager wants to make sure the process is runs flawlessly. For this to happen, the plant manager knows that a running pump will evetually fail, but he would like to know when this is going to happen so he/she can schedule the maintenace. 

Over the years, pumps have become more and more efficient in their operation but equaly expensive to purchase and maintain. To enhance their efficiency, pumps are installed as near to the source of what they are hauling as poosible. Often, as in the case of an oil well, this would be thousands of feet below the earth. To maintain the pumps, an engineer would have to extract the pump, assess the failure, repair, and reinstall. 

*Impact of Failing Pumps*

When a pump fails, it almost always stalls an entire process, as liquids are not moved from one location to the other thus causing pile up  on both the loading and the receiving silos. The damage may be as extensive and destructive. For the oil well, this could mean rendering idle heavy machinery and field workers as they await replacement. This unscheduled maintenance may end up costing the organization millions of dollars in stalled equipment (if under lease), idle labor costs, and revenue losses. In addition, pump replacement requires specialized technicians and in some instances takes severall hours to uninstall the bad pump and install the new/repaired pump. One pump failure could also result in a cascade effect causing failure in multiple processes, and costing thousands of dollars if it happens without warning and proper preparations. For example, it may require making immediate arrangements to divert oil delivery to other reservoirs to minimize losses, or stalling the escavation, managing the receiving units so they don't break down as well, or doing whatever possible to minimize losses from a stalled pump.

*Causes of Pump Failure*

Pump failure can happen anytime for a myriad reasons. In an article appearing in the Plant Wellness Way EAM website, *List of 600+ Centrifugal Pump Set Failure Modes* indicated there are over 600 *pump* failure modes. In other words, unless you see the actual failed unit it may be difficult to know why the unit failed.This is worrisome to any plant manager interested in meeting deadlines and avoiding pump failures. The article distributes pump failure between human error e.g. installation error, wrong pump being installed, etc and Materials-of-Construction failure i.e. equipment failure^[Https://www.lifetime-reliability.com/cms/free-articles/reliability-improvement/pump-failure-modes-list/]. With so much at stake, monitoring the pumps becomes a top priority. Organizations, such as Flowserve, are now using artficial intelligence to implement solutions that will improve efficiency and minimize losses.

### PROBLEM STATEMENT
Oil pumps are critical to the efficiency of oil wells. When an oil well is dug, a pump is required to pump the oil out of the well into a storage or to a delivery pipe that delivers the oil into a reservoir, where another pump will be used to pump the oil into the refinery. When the oil is refined, it will require another pump to pump the oil into other reservoirs waiting distribution. When a (one) pump therefore fails, it's not only costly to remove and replace it, but there is also a cascade effect as it involves not only halting oil production, but transportation as well, jamming reservoirs, and utilizing specialized personnel and equipment to pull the pump out of the well and replace it with a new one according to an article "Pump Failure Prediction."^[Https://www.maana.io/digital-transformation/pump-failure-prediction/] 

If we can successfuly predict pump failure before it happens, we can provide this information to the maintenance team who will in turn schedule a shut down, and replace all the pumps predicted as calpable to failure. This scheduled maintenance will reduce the indiual replacement cost as specialized pesonnel will pull and replace several pumps; In addition, the prior arrangement will have been made to manage other dependant reservoirs, and also send staff on vacation as the maintenance is being performed.

This case study will, based on available data sets, apply Data Science to predict pump failure and help clients avoid closing their plans other than through a scheduled maintenance process.

### DATA SETS AND APPROACH
Our client, a pump manufacturer, provided us with eight data sets i.e BearingWear$Outcome; Cavitation$Outcome; ClosedPressure$Outcome; ClosedSuction$Outcome;CouplingWear$Outcome; Imbalance$Outcome; and Misalignment$Outcome. These data sets were obtained from **factory standards** which we considered to be relatively clean. Our client also provided us with similar data sets obtained from the field.
Using these data sets, we shall use R to read the data sets, modify them where necessary and build a model. We shall then use RandomForest on this model to predict pump failure using the categories afore listed. 


```{r}
# read and clean the data
BearingWear <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Bearing_Wear.csv")
head(BearingWear)
Cavitation <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Cavitation.csv")
ClosedPressure <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Closed_pressure_side.csv")
ClosedSuction <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Closed_Suction_Side.csv")
CouplingWear <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Coupling_Wear.csv")
Imbalance <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Imbalance.csv")
Misalignment <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Misalignment.csv")
Normal_Ops <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Normal_Operation.csv")

BearingWear$Outcome <- c("Bearing Wear")
Cavitation$Outcome <- c("Cavitation")
ClosedPressure$Outcome <- c("ClosedPressure")
ClosedSuction$Outcome <- c("ClosedSuction")
CouplingWear$Outcome <- c("CouplingWear")
Imbalance$Outcome <- c("Imbalance")
Misalignment$Outcome <- c("Misalignment")
Normal_Ops <- c("Normal")

clean_data <- merge(BearingWear, Cavitation)
head(clean_data)
head(Cavitation)
clean_data <- rbind(BearingWear, ClosedPressure)
```


### Read the data from file
```{r}
# Install the required library
library(randomForest)
library(ggplot2)
# Read the data from file
# Following data is from a pump in the lab. Different conditions were induced and data was captured.
BearingWear <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Bearing_Wear.csv")
head(BearingWear)
Cavitation <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Cavitation.csv")
ClosedPressure <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Closed_pressure_side.csv")
ClosedSuction <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Closed_Suction_Side.csv")
CouplingWear <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Coupling_Wear.csv")
Imbalance <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Imbalance.csv")
Misalignment <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Misalignment.csv")
Normal_Ops <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Normal_Operation.csv")

# Additional data received from the pumps in the field
field_cav <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Cavitation_2.csv")
field_coup <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Coupling_Wear_4.csv")
field_mis <- read.csv("C:\\Users\\KRamasundaram\\Documents\\MSDS 6306\\Case Study 2\\Data Set\\Misalignment_2.csv")
```

### Added a column for outcome
```{r}
BearingWear$Outcome <- c("Bearing Wear")
Cavitation$Outcome <- c("Cavitation")
ClosedPressure$Outcome <- c("Closed Pressure Side")
ClosedSuction$Outcome <- c("Closed Suction Side")
CouplingWear$Outcome <- c("Coupling Wear")
Imbalance$Outcome <- c("Imbalance")
Misalignment$Outcome <- c("Misalignment")
Normal_Ops$Outcome <- c("Normal")
field_cav$Outcome <- c("Cavitation")
field_coup$Outcome <- c("Coupling Wear")
field_mis$Outcome <- c("Misalignment")
BearingWear$Outcome <- c(1)
Cavitation$Outcome <- c(2)
ClosedPressure$Outcome <- c(3)
ClosedSuction$Outcome <- c(4)
CouplingWear$Outcome <- c(5)
Imbalance$Outcome <- c(6)
Misalignment$Outcome <- c(7)
Normal_Ops$Outcome <- c(8)
field_cav$Outcome <- c(2)
field_mis$Outcome <- c(7)
field_coup <- field_coup[, -c(38:68)]
field_coup$Outcome <- c(5)

```

### Splitting the data for validation and model building
```{r}
Bearing_val_data <- BearingWear[c(1:as.integer(0.2*nrow(BearingWear))) ,]
Cavitation_val_data <- Cavitation[c(1:as.integer(0.2*nrow(Cavitation))) ,]
ClosedPressure_val_data <- ClosedPressure[c(1:as.integer(0.2*nrow(ClosedPressure))) ,]
ClosedSuction_val_data <- ClosedSuction[c(1:as.integer(0.2*nrow(ClosedSuction))) ,]
CouplingWear_val_data <- CouplingWear[c(1:as.integer(0.2*nrow(CouplingWear))) ,]
Imbalance_val_data <- Imbalance[c(1:as.integer(0.2*nrow(Imbalance))) ,]
Misalignment_val_data <- Misalignment[c(1:as.integer(0.2*nrow(Misalignment))) ,]
Normal_val_data <- Normal_Ops[c(1:as.integer(0.2*nrow(Normal_Ops))) ,]
```

### combine all validation data
```{r}
validation_data <- rbind (Bearing_val_data, Cavitation_val_data)
validation_data <- rbind(validation_data, ClosedPressure_val_data)
validation_data <- rbind(validation_data, ClosedSuction_val_data)
validation_data <- rbind(validation_data, CouplingWear_val_data)
validation_data <- rbind(validation_data, Imbalance_val_data)
validation_data <- rbind(validation_data, Misalignment_val_data)
validation_data <- rbind(validation_data, Normal_val_data)

dim(validation_data)
```

### Model data
```{r}
Bearing_mod_data <- BearingWear[c(1:as.integer(0.8*nrow(BearingWear))) ,]
Cavitation_mod_data <- Cavitation[c(1:as.integer(0.8*nrow(Cavitation))) ,]
ClosedPressure_mod_data <- ClosedPressure[c(1:as.integer(0.8*nrow(ClosedPressure))) ,]
ClosedSuction_mod_data <- ClosedSuction[c(1:as.integer(0.8*nrow(ClosedSuction))) ,]
CouplingWear_mod_data <- CouplingWear[c(1:as.integer(0.8*nrow(CouplingWear))) ,]
Imbalance_mod_data <- Imbalance[c(1:as.integer(0.8*nrow(Imbalance))) ,]
Misalignment_mod_data <- Misalignment[c(1:as.integer(0.8*nrow(Misalignment))) ,]
Normal_mod_data <- Normal_Ops[c(1:as.integer(0.8*nrow(Normal_Ops))) ,]
```

### Consolidate All Model Data
```{r}
model_data <- rbind (Bearing_mod_data, Cavitation_mod_data)
model_data <- rbind(model_data, ClosedPressure_mod_data)
model_data <- rbind(model_data, ClosedSuction_mod_data)
model_data <- rbind(model_data, CouplingWear_mod_data)
model_data <- rbind(model_data, Imbalance_mod_data)
model_data <- rbind(model_data, Misalignment_mod_data)
model_data <- rbind(model_data, Normal_mod_data)
# Make the colunn names in the field data set match training data set
colnames(field_coup) <- colnames(validation_data)
dim(model_data)
dim(validation_data)

```

### Training and Testing
```{r}

# Build the model using minimum number of sensors and exclude the outcomes
M <- model_data[, c(-1, -2, -c(3:16), -38)] 
L <- as.factor(model_data[, 38]) #labels, so select only target i.e. outcome

# trainig and testing
pump.rf <- randomForest(M, L, ntree = 10)

# evaluating the model
pump.rf
pump.rf$importance
pred <- as.data.frame(predict(pump.rf, validation_data[, c(-1, -2, -c(3:16) -38)]))
colnames(pred) <- c("Outcome")
```

### plot for cross-validation testing
```{r}
pump_state <- c("Bearing Wear", "Cavitation", "Closed Pressure", "Clsoed Suction", "Coupling Wear", "Imbalnce", "Misalignment", "Normal")
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred$Outcome), col = 'red', main = "Predicted Outcomes", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
barplot(table(validation_data[, 38]), col = 'blue', main = "Actual Outcomes", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
```

### Anamoly Detection & Prediction of Cavitation Failure on field data
```{r}
# predicitng caviation failure for the pump from the field
pred_cav <- as.data.frame(predict(pump.rf, field_cav[c(2:14000), c(-1, -2, -(3:16),-38)]))
colnames(pred_cav) <- c("Outcome")

# Anamoly Detector for the data from the field
anamoly <- c("Normal", "Anamoly")
pred_cav$anamoly <- ifelse(as.numeric(pred_cav$Outcome)<8, 1, 0)

# plotting prediction
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_cav$Outcome), col = 'red', main = "Predicted Outcomes Using Lab Data For Cavitation", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
# Count of anamaly prediction 
table(pred_cav$anamoly)
barplot(table(pred_cav$anamoly), col = 'green', main = "Anamoly Detector for Cavitation Data", names.arg = anamoly,  horiz = TRUE, las = 1)
```

### Prediction of Cavitation after Adding 20% of the data from the field
```{r}
model_data <- rbind(model_data, field_cav[c(1:as.integer(0.2*nrow(field_cav))) ,])
M <- model_data[, c(-1, -2, -c(3:16), -38)] 
L <- as.factor(model_data[, 38]) #labels, so select only target i.e. outcome

# re-trainig
pump.rf <- randomForest(M, L, ntree = 10)


# predicting again
pred_cav <- as.data.frame(predict(pump.rf, field_cav[c(2:14000), c(-1, -2, -(3:16),-38)]))
colnames(pred_cav) <- c("Outcome")

# plottting the prediction
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_cav$Outcome), col = 'red', main = "Predicted Outcomes Adding 20% of Field Data to Model for Cavitation", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)

```

### Anamoly Detection & Prediction of Misalignment on field data
```{r}
# predicting misalignment failure
pred_misalignment <- as.data.frame(predict(pump.rf, field_mis[c(2:820), c(-1, -2, -(3:16),-38)]))
colnames(pred_misalignment) <- c("Outcome")
pred_misalignment$anamoly <- ifelse(as.numeric(pred_misalignment$Outcome)<8, 1, 0)


# plotting prediction
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_misalignment$Outcome), col = 'red', main = "Predicted Outcomes Using Lab Data For Misalignment", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
# Count of anamaly detector
table(pred_misalignment$anamoly)
barplot(table(pred_misalignment$anamoly), col = 'green', main = "Anamoly Detector for Misalignment Data", names.arg = anamoly,  horiz = TRUE, las = 1)

```

### Prediction of Misalignment after Adding 20% of the data from the field
```{r}
# Adding 20% of the data from the field to improve accuracy
model_data <- rbind(model_data, field_mis[c(1:as.integer(0.2*nrow(field_mis))) ,])
M <- model_data[, c(-1, -2, -c(3:16), -38)] 
L <- as.factor(model_data[, 38]) #labels, so select only target i.e. outcome

# trainig and testing
pump.rf <- randomForest(M, L, ntree = 10)

# predicitng misalginment for the data from the field
pred_misalignment <- as.data.frame(predict(pump.rf, field_mis[c(2:820), c(-1, -2, -(3:16),-38)]))
colnames(pred_misalignment) <- c("Outcome")


#plotting the predicition
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_misalignment$Outcome), col = 'red', main = "Predicted Outcomes After Adding 20% Field Data For Misalignment", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
```

### Anamoly Detection & Prediction of coupling failure on field data
```{r}
# predicitng coupling failure
# Data clean up
field_coup[, c(17:37)] <- lapply(field_coup[, c(17:37)], as.character)
field_coup[, c(17:37)] <- lapply(field_coup[, c(17:37)], as.numeric)

# Prediction
pred_coup <- as.data.frame(predict(pump.rf, field_coup[c(2:75000), c(-1, -2, -(3:16),-38)]))
colnames(pred_coup) <- c("Outcome")
pred_coup$anamoly <- ifelse(as.numeric(pred_coup$Outcome)<8, 1, 0)

# plotting prediction
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_coup$Outcome), col = 'red', main = "Predicted Outcomes Using Lab Data For Coupling Failure", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)
# Count of anamaly prediction 
table(pred_coup$anamoly)
barplot(table(pred_coup$anamoly), col = 'green', main = "Anamoly Detector for Coupling Data", names.arg = "Anamoly",  horiz = TRUE, las = 1)

```


### Prediction of coupling after Adding 20% of the data from the field
```{r}

# Adding 20% of the data from the field to improve accuracy
model_data <- rbind(model_data, field_coup[c(1:as.integer(0.2*nrow(field_coup))) ,])

M <- model_data[, c(-1, -2, -c(3:16), -38)] 
L <- as.factor(model_data[, 38]) #labels, so select only target i.e. outcome

# trainig and testing
pump.rf <- randomForest(M, L)

pred_coup <- as.data.frame(predict(pump.rf, field_coup[c(2:75000), c(-1, -2, -(3:16),-38)]))
colnames(pred_coup) <- c("Outcome")

# plotting the prediciton
par(mar = c(5.1, 7.5, 4.1, 2.1))
barplot(table(pred_coup$Outcome), col = 'red', main = "Predicted Outcomes After Adding 20% Field Data For Coupling Failure", xlab = "Occurances", names.arg= pump_state, horiz = TRUE, las = 1)


```



****
[^1]:While we have made references to general pumps including pumps used in the huamn body, this study focuses only on industrial pumps movin liquids. 
[^2]: Https://www.lifetime-reliability.com/cms/free-articles/reliability-improvement/pump-failure-modes-list/
[^3]: Https://www.maana.io/digital-transformation/pump-failure-prediction/